An Efficient Calculus for Generalised Parikh Images of Regular Languages

### 2022-06-22 main.tex
Ignore generated PDFs

\subsubsection{Producing the Parikh image}

Now assume that rather than the length (e.g. the count of the characters) we
want to find the whole Parikh image of the intersection of $\AcaOrBc{}$ and
$\SomethingCSomething{}$. In other words, now $f$ is an object (usually a
vector) counting how often each character occurs in a string. Since extracting
values from a vector is tiresome, we will use the shorthand notation
$\TransitionCount{a}$ to refer to the number of letters a that appear. E.g. $\TransitionCount{a}(\text{curious and curiouser still}) = 1$. Note that we still have the property that $\TransitionCount{o}(\text{homomorphism}) = \TransitionCount{o}(\text{homo}) + \TransitionCount{o}(\text{morphism}) = 2 + 1 = 3$. If you treat the addition as element-wise vector addition, this approach works for the vector version as well.

@@

Thus, we would have $\ParikhMap(\left\{ab, abb\right\}) = \left\{\left[1,
1\right], \left[1, 2\right]\right\}$.

Parikh's theorem states that ignoring order, any context-free grammar has a
letter-equivalent regular language (c.f.~\cite{construction} for a construction of
such automata from context-free grammars and~\cite{bounds} for bounds on its size). However, there are languages that are
not context-free that also have semilinear images under~$\ParikhMap$ (e.g.
$\ParikhMap(\Set{a^nb^nc^n \SuchThat n \geq 0}) = \ParikhMap((abc)^*) = \#a =
\#b = \#c \land \#a \geq 0$).

@@

\section{Parikh Images from Products of Automata}

@@

\section{Extensions}
\subsection{Backjumping and Learning No-Goods}
\subsubsection{An Example}
\subsection{Symbolic Automata}\label{sec:ext:symbolic}

@@

@@

We implement the baseline approach using the same Presburger solver
(\Princess{}), input file parser, and automaton implementation as \Catra{}. We
do this in order to make the comparison between methods as fair as possible, and
give us the ability to compare the efficiency of the calculus rules themselves.
Using the formula of Equation~\ref{eq:generate-parikh}, we produce quantified
Presburger formulae for each successive term and add them to Princess. In this
fashion we compute the product incrementally term by term, checking
satisfiability in each step. We use \Fudge{a priority queue} to select which
term to use for the next step of the product, and order it by \Fudge{the number
of transitions} as a heuristic for the size of the term. We use this heuristic
to avoid computing large (and therefore slow) products until we have to, banking
our hopes on finding a source of unsatisfiability earlier. The pseudocode for
our implementation can be seen in Algorithm~\ref{alg:baseline}.

@@

It should be noted that our automata (including the successive products) are
always by construction forward- and backward- reachable-minimal. We avoid
computing fully minimal automata since it is unclear when an automaton with
registers can be safely minimised. This guarantees that any automaton we produce
only contains states that are both reachable from the initial state and has a
path to an accepting state.

\subsection{Heuristics and search strategies}
\contents{
  \item Optimisations and tricks
}

>>>>

\subsubsection{Producing the Parikh image}\label{sec:introduction:parikh}

Now assume that rather than the length (e.g. the count of the characters) we
want to find the whole Parikh image of the intersection of $\AcaOrBc{}$ and
$\SomethingCSomething{}$. In other words, now $f$ is a mapping to an object
(usually seen as a vector) counting how often each character occurs in a string. Since
extracting values from a vector is tiresome, we will use the shorthand notation
$\TransitionCount{a}$ to refer to the number of letters a that appear. E.g.
$\TransitionCount{a}(\text{curious and curiouser still}) = 1$. Note that we
still have the property that $\TransitionCount{o}(\text{homomorphism}) =
\TransitionCount{o}(\text{homo}) + \TransitionCount{o}(\text{morphism}) = 2 + 1
= 3$. If you treat the addition as element-wise vector addition, this approach
works for the vector version as well.

@@

Thus, we would have $\ParikhMap(\left\{ab, abb\right\}) = \left\{\left[1,
1\right], \left[1, 2\right]\right\}$. We also sometimes use the standard
notation $\#l$ to talk about an individual letter $l$. For example, for the Parikh vector above, we would have $\CountOf{a} = 1$. You have already seen 
this in Section~\ref{sec:introduction:parikh}.

Parikh's theorem states that ignoring order, any context-free grammar has a
letter-equivalent regular language (c.f.~\cite{construction} for a construction of
such automata from context-free grammars and~\cite{bounds} for bounds on its size). However, there are languages that are
not context-free that also have semilinear images under~$\ParikhMap$ (e.g.
$\ParikhMap(\Set{a^nb^nc^n \SuchThat n \geq 0}) = \ParikhMap((abc)^*) = \CountOf{a} =
\CountOf{b} = \CountOf{c} \land \CountOf{a} \geq 0$).

@@

\section{Parikh Images from Products of Automata}\label{sec:multiple}

@@

\section{Extensions}
\subsection{Backjumping and Learning No-Goods}\label{sec:ext:backjumping}
\subsubsection{An Example}
\subsection{Symbolic Automata}\label{sec:ext:symbolic}

@@

\contents{
  \item the unused predicate
  \item single-arity
  \item a princess theory
}

@@

We implement the baseline approach using the same Presburger solver
(\Princess{}), input file parser, and automaton implementation as \Catra. We
do this in order to make the comparison between methods as fair as possible, and
give us the ability to compare the efficiency of the calculus rules themselves.
Using the formula of Equation~\ref{eq:generate-parikh}, we produce quantified
Presburger formulae for each successive term and add them to Princess. In this
fashion we compute the product incrementally term by term, checking
satisfiability in each step. We use \Fudge{a priority queue} to select which
term to use for the next step of the product, and order it by \Fudge{the number
of transitions} as a heuristic for the size of the term. We use this heuristic
to avoid computing large (and therefore slow) products until we have to, banking
our hopes on finding a source of unsatisfiability earlier. The pseudocode for
our implementation can be seen in Algorithm~\ref{alg:baseline}.

@@

It should be noted that our automata (including the successive products) are
always by construction forward- and backward- reachable-minimal. We avoid
computing fully minimal automata since it is unclear when an automaton with
registers can be safely minimised. This guarantees that any automaton we produce
only contains states that are both reachable from the initial state and has a
path to an accepting state. We never perform any other minimisation on the
automata for either backend. More complex minimisation was left out since
performing minimisation on automata with counters is non-trivial, and indeed
already minimising symbolic automata is itself risks exponential blowup
\cite{minimising-symbolic}.

\subsection{Heuristics and search strategies}

There are a number of choices left unspecified in \Calculus{} as described in
Sections~\ref{sec:calculus} and~\ref{sec:multiple}. For example, the order of
materialisation of intermediate products and the order of splitting. In
addition, the splitting specified in the general calculus is correct but
inefficient. In this section we describe additional implementation details used
to enhance our solution.

\subsubsection{Splitting, Materialisation, and Propagation}

We order our rule applications so that we first propagate connectedness if
possible, then perform materialisation if tractable (see below!), then finally
resort to splitting if we must.

In addition to the splitting on individual transition variables described in Table~\ref{tbl:rules:single}, where we randomly select a variable whose bounds are not known, we also prefer splitting to sever/include a strongly connected component. We randomly select an automaton where we can compute a cut to at least one strongly connecting component that can be severed from the initial state, e.g. where the strongly connected component does not contain the initial state and where the sum of the transition variables we have associated with the transitions of a minimal cut between the initial state and an arbitrary state in the component is not known to be greater than 0, but also not equal to 0. If there are multiple such strongly connected components we choose one randomly. We then proceed to split on the cut as if it were a regular transition, e.g. its sum being zero or nonzero. In this way we use our splitting strategy to drive the connectedness constraint towards propagation by cutting off loops from the automata.

The implementation of the connectedness constraint is opportunistic and straightforward. We compute a set of dead states by performing forward and backwards reachability computations on an automaton, where we disregard any transition whose associated variable is known to be zero. After that we add clauses ensuring any transition variable associated with a transition starting in a dead state is zero.

Product materialisation, then, is the final piece of the puzzle. In the current implementation, we put off computing intermediate products until at least all but 2 transition variables of one of them have a known status, e.g. either is known to be zero or known to be positive. The number was chosen experimentally, and we observe a consistent trend that to have a low number and therefore a high threshold for materialisation is better throughout the implementation of other features. The other automaton for the product is selected randomly.

\subsubsection{Clause Learning}

Since \Catra{} does not learn very sophisticated clauses yet (such as the ones
described in Section~\ref{sec:ext:backjumping}), it is turned off by default.
However, even in its underdeveloped state, clause learning has shown to
significantly improve the performance on unsatisfiable instances, at the cost of
a loss in performance for satisfiable instances. Since \Catra{} already
struggles more with satisfiable instances than unsatisfiable ones, we opted to
turn the option off by default.

\subsubsection{Random Restarts}

Finally, we perform restarts scaled by the Luby series~\cite{luby}. Experimental
results have shown this to have a large improvement in performance, which is
unsurprising given how many random choices we make during solving and how
tail-heavy our problem is. It seems to be very easy to get stuck with a poor
choice of product to materialise or transitions to split on.
